---
title: EDA with Categorical Variables
author: David Timewell
date: '2020-09-13'
categories:
  - R
tags:
  - Categorical Variables
  - EDA
slug: index.en-us
keywords: tech
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Intro to EDA with Categorical variables

Below is an introduction to EDA with categorical variables. THis should give a decent starting point when you want to start
your analysis.

If you knowledge of ggplot needs some work then this book is really good : https://r-graphics.org/


```{r warning=FALSE}
 
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(kableExtra)
```

We are going to be using crash data from the QLD government.
So lets go ahead and grab it.

```{r}
 
crash_data <- read_csv("http://www.tmr.qld.gov.au/~/media/aboutus/corpinfo/Open%20data/crash/locations.csv")



```

I am going to trim this down to investigate speed zone and crash severity as my categorical variables. And I am only interested in Motorcycle accidents.


```{r}
 
crash_data_eda <- crash_data %>% filter(Count_Unit_Motorcycle_Moped > 0 )  %>% select(Crash_Severity,Crash_Speed_Limit) 

crash_data_eda <- na.omit(crash_data_eda)
```


Take a look at the structure of the data.

```{r}
glimpse(crash_data_eda)

```

Lets convert the variables to factors.

```{r}

crash_data_eda$Crash_Severity <- as.factor(crash_data_eda$Crash_Severity) 
crash_data_eda$Crash_Speed_Limit <- as.factor(crash_data_eda$Crash_Speed_Limit) 


```

Lets start by looking at the variables by themselves. We will start with the speed zone. 
Looks like 60km/h zone is where most accidents occur.

```{r}

ggplot(crash_data_eda,aes(x=Crash_Speed_Limit)) +
  geom_bar()

```

Now let's pretty it up and make it more meaningful by including titles , labels ect. 
I wil also reorder the factor.


```{r}

crash_data_eda$Crash_Speed_Limit <- factor(crash_data_eda$Crash_Speed_Limit,levels = c("60 km/h","0 - 50 km/h","100 - 110 km/h","80 - 90 km/h","70 km/h"))

ggplot(crash_data_eda,aes(x=Crash_Speed_Limit)) +
  geom_bar(colour="black",fill="#CCEEFF") +
   xlab("Speed Zone") + 
    ylab("Number of Accidents") +
     geom_text(aes(label = ..count..), stat = "count", vjust = 1.5,
            colour = "black") +
       ggtitle("Number of Motorcycle Accidents by Speed Zone","(2001 - 2018)") +
         theme_bw() 

```


We can also look at the counts in a table.

```{r}

options(scipen = 999, digits = 3)   #Simplify display format

 
tab_cnt <- table(crash_data_eda$Crash_Speed_Limit) 
tab_cnt      

```

And now get a table of proportions and save into props to use in the next section

```{r}


(props <- prop.table(tab_cnt))

```

Lets plot this percent of total accidents now for each speed zone.

```{r}
as.data.frame(props) %>% rename("Speed Zone"=Var1,"Percent"=Freq) %>% mutate(percent_lab = paste(as.character(round(Percent*100)),"%")) %>% ggplot(aes(x=`Speed Zone`,y=Percent)) +
  geom_bar(stat='identity',colour="black",fill="#CCEEFF") +
   scale_y_continuous(labels = scales::percent) +
    theme_bw() +
      geom_text(aes(label = percent_lab,vjust = 1.5 )) +
           ggtitle("Proportion of Motorcycle Accidents by Speed Zone","(2001 - 2018)")

```


You can try the same with the other variable, but lets now start looking at the combination of Crash_Speed_Limit and Crash_Severity.

Again we can use the table to take a look at counts.

```{r}

(tab_cnt <- table(crash_data_eda$Crash_Speed_Limit, crash_data_eda$Crash_Severity))


```


I am going to trim down my data set to Hospitalisation vs Fatal as these make up the greatest proportion.

```{r}

crash_data_eda <- crash_data_eda %>% filter(!Crash_Severity %in%  c("Property damage only","Minor injury","Medical treatment")) 



```


If we look at counts again notice the factor levels are still represented but are zero. Note I used kableExtra just to show a different formatting option. There is alot more to the package take a look at :
https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

```{r}

(tab_cnt <- table(crash_data_eda$Crash_Speed_Limit, crash_data_eda$Crash_Severity))  %>% kbl() %>%   kable_paper(bootstrap_options = "striped",full_width = T,position = "float_left") %>% footnote(general = "Crash frequency by speed zone (2001-2018). ")



```


We can drop levels to clean things up.

```{r}

crash_data_eda$Crash_Severity <- droplevels(crash_data_eda$Crash_Severity)
tab_cnt <- table(crash_data_eda$Crash_Speed_Limit, crash_data_eda$Crash_Severity)

```



Lets look at proportions

```{r}

(prop.table(tab_cnt)) 


```


If we want proportion by rows (rows sum to 1)

```{r}

(prop.table(tab_cnt,1))
```

If we want proportion by columns (columns sum to 1). We can also round and mutiply by 100 to make it look more like percents.

```{r}



  round(prop.table(tab_cnt,2) * 100)


```

Lets plot some of these. And start with plotting counts  



```{r fig.width=15}

ggplot(crash_data_eda, aes(x = Crash_Speed_Limit, fill = Crash_Severity)) + 
  geom_bar(position = "dodge",colour="black") +
   scale_fill_manual(values = c("#FFDDDD","#CCEEFF")) +
     xlab("Speed Zone") + 
       ylab("Number of Accidents") +
          guides(fill=guide_legend(title=NULL)) +
            ggtitle("Number of Fatal vs Hospitalisation Motorcycle Accidents by Speed Zone","(2001 - 2018)") +
              theme_bw() +
                geom_text(aes(label = ..count..), stat = "count", vjust = -0.2,hjust= 1,
                colour = "black")
            

```


Lets plot the proportions , it looks like 60km and 100-110km zones make up the highest proportion of fatalities and hospitalisations. 

```{r }

 
plot_data_eda <- crash_data_eda %>% group_by(Crash_Severity)  %>% count(Crash_Speed_Limit) %>% mutate(pct= prop.table(n) )  %>% mutate(pct_lab = as.character(round(pct*100))) %>% arrange(Crash_Speed_Limit,desc(pct))



ggplot(plot_data_eda,aes(x=Crash_Severity,y=pct,fill=Crash_Speed_Limit)) +     
   geom_col(colour="black") +
     geom_text(position = position_stack(vjust = .5),aes(x=Crash_Severity,y=pct,label=   paste(pct_lab,"%")   )) +
     scale_y_continuous(labels = scales::percent) + 
       scale_fill_brewer(palette = "Pastel1")   +
           guides(fill=guide_legend(title=NULL)) +
             xlab("Crash Severity Type") + 
               ylab("") + 
               ggtitle("Proportion of Motorcycle Accidents by Speed Zone","(2001 - 2018)") +
                  theme_bw() 



```



Now lets look at the proportion of crash severity in each speed zone. 
Looks like 100-110km has the highest proportion of fatalities. I think speed could be a factor here.

```{r}
plot2_data_eda <- crash_data_eda %>% group_by(Crash_Speed_Limit)  %>% count(Crash_Speed_Limit,Crash_Severity) %>% mutate(pct= prop.table(n) )  %>% mutate(pct_lab = as.character(round(pct*100))) %>% arrange(Crash_Speed_Limit,desc(pct))



plot2_data_eda$Crash_Speed_Limit <- factor(plot2_data_eda$Crash_Speed_Limit,levels = c("100 - 110 km/h","80 - 90 km/h","70 km/h","60 km/h","0 - 50 km/h"))

ggplot(plot2_data_eda,aes(x=Crash_Speed_Limit,y=pct,fill=Crash_Severity)) +     
   geom_col(colour="black") +
       scale_fill_manual(values = c("#FFDDDD","#CCEEFF"))   +
           guides(fill=guide_legend(title=NULL)) +
         xlab("Speed Zone") + 
           ylab("") + 
               ggtitle("Proportion of Motorcycle Accidents by Speed Zone","(2001 - 2018)") +
                  theme_bw() +
                    geom_text(position = position_stack(vjust = .5),aes(x=Crash_Speed_Limit,y=pct,label=   paste(pct_lab,"%")   )) + 
                       scale_y_continuous(labels = scales::percent) 



```


So there we go, some basic charts to use and tables to investigate categorical variables as a start to getting some insight.
Also take a look  at :
For a look into contingency tables :
https://medium.com/@nhan.tran/contingency-table-55e375a76a8
And for Chi Square :
https://medium.com/@nhan.tran/the-chi-square-statistic-p-1-37a8eb2f27bb


